---
title: "QSM_OEF_Aug18_JCBFM"
author: "Ahmed Khalil"
date: "13 August 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(lme4)
require(lmerTest)
require(effects)
library(emmeans)
library(kableExtra)
library(MuMIn)
```

**Methods**  
Linear mixed models were used to investigate:

1. the main effects of hemisphere (contralateral / ischemic) and group (no hypoperfusion / match / large mismatch), as well as their interaction, on the OEF values using the R package "lme4" [1]. The model included varying intercepts for each patient and hemisphere.  
2. the relationship between OEF and rCBF in the ischemic hemispheres (with varying intercepts for each patient).  
3. the relationship between the distance from the DWI lesion and the OEF value. This included the pattern of hypoperfusion (match vs large mismatch) as a fixed effect (with varying intercepts for each patient).  
4. the relationship between the ineraction of NIHSS and time (baseline / follow up) with the OEF ratio (with varying intercepts for each patient).

P-values of the mixed models' coefficients were calculated using Satterthwaite's method for approximating degrees-of-freedom (R package “lmerTest” [2]). A visualization of the models was created using the R package "effects" [3]. Marginal and conditional R-squared were calculated using the R package "MuMIn" [4].

*References:*  
[1] Douglas Bates, Martin Maechler, Ben Bolker, Steve Walker (2015). Fitting Linear Mixed-Effects Models Using lme4.
  Journal of Statistical Software, 67(1), 1-48. doi:10.18637/jss.v067.i01.  
  
[2] Kuznetsova A, Brockhoff PB and Christensen RHB (2017). “lmerTest Package: Tests in Linear Mixed Effects Models.”
Journal of Statistical Software, 82(13), pp. 1-26. doi: 10.18637/jss.v082.i13 (URL:
http://doi.org/10.18637/jss.v082.i13).  

[3] John Fox (2003). Effect Displays in R for Generalised Linear Models. Journal of Statistical Software, 8(15), 1-27.
  URL http://www.jstatsoft.org/v08/i15/.  

[4]  Kamil Barton (2018). MuMIn: Multi-Model Inference. R package version 1.42.1. https://CRAN.R-project.org/package=MuMIn
  
# Model 1
```{r, echo= FALSE, fig.width=4, fig.height=4, fig.align='center',dpi=300, warning=F}
# load OEF data
OEF_data <- read.csv(file = "OEFs_AK_mm.csv", header = T, sep=";")
# re-order groups so that "no hypoperfusion" is the reference
OEF_data$Group <- factor(OEF_data$Group, levels=c("no_HP", "match", "large_MM"))
levels(OEF_data$Group) <- c("No hypoperfusion", "Match", "Large mismatch")
# build mixed model
OEF_model <- lmer(formula = OEF~Group*Hemisphere + (Hemisphere|Patient), data = OEF_data)
# mixed model ANOVA
anova(OEF_model)
# mixed model results
summary(OEF_model)
# get R^2 of mixed model
OEF_rsq <- r.squaredGLMM(OEF_model)
# prepare for effects visualization
OEF_effects <- allEffects(OEF_model)
# plot effects
plot(OEF_effects, multiline=T, ci.style="bars", main="", xlab="")
```

# Model 2
```{r, echo= FALSE, fig.width=4, fig.height=4, fig.align='center',dpi=300, warning=F}
# load OEF data
OEF_CBF_data <- read.csv(file = "OEF_CBF_AK_full_mm.csv", header = T, sep = ",")
# re-order groups so that "no hypoperfusion" is the reference
OEF_CBF_data$Group <- factor(OEF_CBF_data$Group, levels=c("no_HP", "match", "large_MM"))
# build mixed model
OEF_CBF_model <- lmer(formula = OEF~rCBF*Group + (1|Patient), data=OEF_CBF_data)
# mixed model ANOVA
anova(OEF_CBF_model)
# mixed model results
summary(OEF_CBF_model)
# get R^2 of mixed model
OEF_CBF_rsq <- r.squaredGLMM(OEF_CBF_model)
# prepare for effects visualization
OEF_CBF_effects <- allEffects(OEF_CBF_model)
# plot effects
plot(OEF_CBF_effects,main="")

### SAME AS ABOVE, for relative OEF ### 
OEF_CBF_model_rel <- lmer(formula = rOEF~rCBF*Group + (1|Patient), data=OEF_CBF_data)
anova(OEF_CBF_model_rel)
summary(OEF_CBF_model_rel)
OEF_CBF_rsq_rel <- r.squaredGLMM(OEF_CBF_model_rel)
OEF_CBF_effects_rel <- allEffects(OEF_CBF_model_rel)
plot(OEF_CBF_effects_rel,main="")
```

# Model 3
```{r, echo= FALSE, fig.width=4, fig.height=4, fig.align='center',dpi=300, warning=F}
# load OEF data
OEF_dist_data <- read.csv(file = "OEFs_Distance_AK_mm.csv", header = T, sep = ";")
# re-order groups so that "match" is the reference
OEF_dist_data$Group <- factor(OEF_dist_data$Group, levels = c("Match", "Large mismatch"))
# build mixed model
OEF_dist_model <- lmer(formula = OEF~Distance*Group + (1|Patient), data=OEF_dist_data)
# mixed model ANOVA
anova(OEF_dist_model)
# mixed model results
summary(OEF_dist_model)
# get R^2 of mixed model
OEF_dist_rsq <- r.squaredGLMM(OEF_dist_model)
print(OEF_dist_rsq)
# prepare for effects visualization
OEF_dist_effects <- allEffects(OEF_dist_model)
# plot effects
plot(OEF_dist_effects,main="")
```
# Model 4
```{r, echo= FALSE, fig.width=4, fig.height=4, fig.align='center',dpi=300, warning=F}
# load OEF data
OEF_fu_data <- read.csv(file = "OEF_NIHSS_mRS_followup_AK_mm.csv", header = T, sep = ",")
# build mixed model
OEF_fu_model <- lmer(formula = OEF_ratio~Timepoint*NIHSS + (1|Patient), data=OEF_fu_data)
# mixed model ANOVA
anova(OEF_fu_model)
# mixed model results
summary(OEF_fu_model)
# get R^2 of mixed model
OEF_fu_rsq <- r.squaredGLMM(OEF_fu_model)
print(OEF_fu_rsq)
# prepare for effects visualization
OEF_fu_effects <- allEffects(OEF_fu_model)
# plot effects
plot(OEF_fu_effects,main="")
```

**Results**  
There was no statistically significant main effect of either group (F = 3.18, p = 0.07) or hemisphere (F= 3.98, p = 0.05) on OEF values. The marginal R-squared (fixed effects only) was `r round(OEF_rsq[1],2)` and the conditional R-squared (fixed + random effects) was `r round(OEF_rsq[2],2)`. There was a statistically significant interaction between group and hemisphere (F = 5.17, p = 0.008), with the difference in OEF between hemispheres being most prominent in patients with large mismatch (b = 0.04, standard error = 0.018, t = 2.2, p = 0.0315).

There was a statistically significant interaction between the distance from the DWI lesion to the vein and the perfusion pattern (match / large mismatch) on the OEF values (b =  -0.0056, standard error = 0.0012, t = -4.54, p < 0.0001). The marginal R-squared (fixed effects only) was `r round(OEF_dist_rsq[1],2)` and the conditional R-squared (fixed + random effects) was `r round(OEF_dist_rsq[2],2)`. 
