---
title: "QSM_OEF_Aug18_JCBFM"
author: "Ahmed Khalil"
date: "13 August 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(lme4)
require(lmerTest)
require(effects)
library(emmeans)
library(kableExtra)
library(MuMIn)
```

**Methods**  
Linear mixed models were used to investigate:

* the main effects of hemisphere (contralateral / ischemic) and group (no hypoperfusion / match / large mismatch), as well as their interaction, on the OEF values using the R package "lme4" [1]. The model included varying intercepts for each patient and hemisphere.  
* the relationship between OEF and rCBF in the ischemic hemispheres of patients with large mismatch (with varying intercepts for each patient).  
* the relationship between the distance from the DWI lesion and the OEF value. This included the pattern of hypoperfusion (match vs large mismatch) as a fixed effect (with varying intercepts for each patient).  

P-values of the mixed models' coefficients were calculated using Satterthwaite's method for approximating degrees-of-freedom (R package “lmerTest” [2]). A visualization of the models was created using the R package "effects" [3]. Marginal and conditional R-squared were calculated using the R package "MuMIn" [4].

*References:*  
[1] Douglas Bates, Martin Maechler, Ben Bolker, Steve Walker (2015). Fitting Linear Mixed-Effects Models Using lme4.
  Journal of Statistical Software, 67(1), 1-48. doi:10.18637/jss.v067.i01.  
  
[2] Kuznetsova A, Brockhoff PB and Christensen RHB (2017). “lmerTest Package: Tests in Linear Mixed Effects Models.”
Journal of Statistical Software, 82(13), pp. 1-26. doi: 10.18637/jss.v082.i13 (URL:
http://doi.org/10.18637/jss.v082.i13).  

[3] John Fox (2003). Effect Displays in R for Generalised Linear Models. Journal of Statistical Software, 8(15), 1-27.
  URL http://www.jstatsoft.org/v08/i15/.  

[4]  Kamil Barton (2018). MuMIn: Multi-Model Inference. R package version 1.42.1. https://CRAN.R-project.org/package=MuMIn
  

```{r, echo= FALSE, fig.width=4, fig.height=4, fig.align='center',dpi=300, warning=F}
# load OEF data
OEF_data <- read.csv(file = "OEFs_AK_mm.csv", header = T, sep=";")
# re-order groups so that "no hypoperfusion" is the reference
OEF_data$Group <- factor(OEF_data$Group, levels=c("no_HP", "match", "large_MM"))
levels(OEF_data$Group) <- c("No hypoperfusion", "Match", "Large mismatch")
# build mixed model
OEF_model <- lmer(formula = OEF~Group*Hemisphere + (Hemisphere|Patient), data = OEF_data)
# mixed model ANOVA
anova(OEF_model)
# mixed model results
summary(OEF_model)
# get R^2 of mixed model
OEF_rsq <- r.squaredGLMM(OEF_model)
# prepare for effects visualization
OEF_effects <- allEffects(OEF_model)
# plot effects
plot(OEF_effects, multiline=T, ci.style="bars", main="", xlab="")
```

```{r, echo= FALSE, fig.width=4, fig.height=4, fig.align='center',dpi=300, warning=F}
# load OEF data
OEF_CBF_data <- read.csv(file = "OEF_CBF_AK_mm.csv", header = T, sep = ";")
# build mixed model
OEF_CBF_model <- lmer(formula = OEF~rCBF + (1|Patient), data=OEF_CBF_data)
# mixed model ANOVA
anova(OEF_CBF_model)
# mixed model results
summary(OEF_CBF_model)
# get R^2 of mixed model
OEF_CBF_rsq <- r.squaredGLMM(OEF_CBF_model)
# prepare for effects visualization
OEF_CBF_effects <- allEffects(OEF_CBF_model)
# plot effects
plot(OEF_CBF_effects,main="")

### SAME AS ABOVE, for relative OEF ### 
OEF_CBF_model_rel <- lmer(formula = rOEF~rCBF + (1|Patient), data=OEF_CBF_data)
anova(OEF_CBF_model_rel)
summary(OEF_CBF_model_rel)
OEF_CBF_rsq_rel <- r.squaredGLMM(OEF_CBF_model_rel)
OEF_CBF_effects_rel <- allEffects(OEF_CBF_model_rel)
plot(OEF_CBF_effects_rel,main="")
```


```{r, echo= FALSE, fig.width=4, fig.height=4, fig.align='center',dpi=300, warning=F}
# load OEF data
OEF_dist_data <- read.csv(file = "OEFs_Distance_AK_mm.csv", header = T, sep = ";")
# re-order groups so that "match" is the reference
OEF_dist_data$Group <- factor(OEF_dist_data$Group, levels = c("Match", "Large mismatch"))
# build mixed model
OEF_dist_model <- lmer(formula = OEF~Distance*Group + (1|Patient), data=OEF_dist_data)
# mixed model ANOVA
anova(OEF_dist_model)
# mixed model results
summary(OEF_dist_model)
# get R^2 of mixed model
OEF_dist_rsq <- r.squaredGLMM(OEF_dist_model)
# prepare for effects visualization
OEF_dist_effects <- allEffects(OEF_dist_model)
# plot effects
plot(OEF_dist_effects,main="")
```
**Results**  
There was no statistically significant main effect of either group (F = 3.18, p = 0.07) or hemisphere (F= 3.98, p = 0.05) on OEF values. The marginal R-squared (fixed effects only) was `r round(OEF_rsq[1],2)` and the conditional R-squared (fixed + random effects) was `r round(OEF_rsq[2],2)`. There was a statistically significant interaction between group and hemisphere (F = 5.17, p = 0.008), with the difference in OEF between hemispheres being most prominent in patients with large mismatch (b = 0.04, standard error = 0.018, t = 2.2, p = 0.0315).

There was a statistically significant relationship between rCBF and OEF (b = -0.22, standard error = 0.05, t = -4.74, p = 0.00008). The marginal R-squared (fixed effects only) was `r round(OEF_CBF_rsq[1],2)` and the conditional R-squared (fixed + random effects) was `r round(OEF_CBF_rsq[2],2)`. 

There was a statistically significant relationship between rCBF and rOEF (b = -0.9, standard error = 0.21, t = -4.34, p = 0.00016). The marginal R-squared (fixed effects only) was `r round(OEF_CBF_rsq_rel[1],2)` and the conditional R-squared (fixed + random effects) was `r round(OEF_CBF_rsq_rel[2],2)`. 

There was a statistically significant interaction between the distance from the DWI lesion to the vein and the perfusion pattern (match / large mismatch) on the OEF values (b =  -0.0056, standard error = 0.0012, t = -4.54, p < 0.0001). The marginal R-squared (fixed effects only) was `r round(OEF_dist_rsq[1],2)` and the conditional R-squared (fixed + random effects) was `r round(OEF_dist_rsq[2],2)`. 
